# Introduction {#intro}

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyr)
library(ggplot2)
library(readr)
library(dplyr)
```
# Environmental Setting

## Climate

The Arctic Coastal Plain of Alaska is characterized by a harsh tundra climate, classified as ET in the Köppen system. This region experiences extreme temperature variations, with brutally cold winters where average temperatures often plummet below -15°F, and brief, cool summers barely reaching 50°F in July. The average annual temperature ranges from 8 to 14°F. The average freeze-free period is fewer than 5 days to 20 days. Freezing temperatures can occur in any month.

```{r, fig.cap = "This is NRCS Snotel data from https://wcc.sc.egov.usda.gov/nwcc/site?sitenum=1177", echo = FALSE, warning=FALSE, message=FALSE}

library(plotly)
library(tidyr)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyverse)
library(lubridate)

df1 <- read_delim("_book/dailyAvgTemp.csv", delim = "|")

data <- df1

data_long <- data %>%
  pivot_longer(cols = -c(`Water Year`, Day),
               names_to = "Month",
               values_to = "Temperature")

month_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

data_long <- data_long %>%
  mutate(
    Month = factor(Month, levels = month_order, ordered = TRUE),
    Date = as.Date(paste("2000", Month, Day, sep = "-"), format = "%Y-%b-%d"),
    `Water Year` = as.factor(`Water Year`),
    hover_text = paste("Water Year:", `Water Year`, "<br>",
                       "Month:", Month, "<br>",
                       "Day:", Day, "<br>",
                       "Temperature:", round(Temperature, 1), "°F")
  ) %>%
  arrange(Month, Day)

# Calculate the average temperature for each day of the year
avg_temp <- data_long %>%
  group_by(Month, Day) %>%
  summarise(AvgTemperature = mean(Temperature, na.rm = TRUE), .groups = 'drop') %>%
  mutate(
    Date = as.Date(paste("2000", Month, Day, sep = "-"), format = "%Y-%b-%d"),
    hover_text = paste("Average Temperature", "<br>",
                       "Month:", Month, "<br>",
                       "Day:", Day, "<br>",
                       "Temperature:", round(AvgTemperature, 1), "°F")
  )

# Create the combined line graph
p <- ggplot() +
  # Individual year lines
  geom_line(data = data_long, 
            aes(x = Date, y = Temperature, color = `Water Year`, group = `Water Year`, text = hover_text), 
            alpha = 0.5) +
  # Average temperature line
  geom_line(data = avg_temp, 
            aes(x = Date, y = AvgTemperature),
            color = "black", size = 1.2) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  scale_color_viridis_d(option = "plasma") +  # Color-blind friendly palette
  labs(title = "Average Daily Temperature 2004-2024",
       x = "Month",
       y = "Temperature (°F)",
       color = "Water Year") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Convert to interactive plotly object with custom hover
p_interactive <- ggplotly(p, tooltip = "text")

# Modify hover template and mode
for (i in 1:length(p_interactive$x$data)) {
  p_interactive$x$data[[i]]$hoverinfo <- "text"
  p_interactive$x$data[[i]]$hoverlabel <- list(bgcolor = "white")
}

p_interactive <- p_interactive %>% 
  layout(hovermode = "closest")

# Display the plot
p_interactive

```

Despite its frigid nature, the arctic coastal plain receives surprisingly little precipitation, typically ranging from 6-10 inches annually. Most of this falls as snow, blanketing the landscape for much of the year. The average annual snowfall is about 50 to 100 centimeters. The low precipitation, coupled with minimal evaporation rates due to the cold, creates a unique hydrologic balance that doesn't quite fit the traditional definition of a desert.

```{r,fig.cap = "This is NRCS Snotel data from https://wcc.sc.egov.usda.gov/nwcc/site?sitenum=1177", echo = FALSE, warning=FALSE, message=FALSE}

library(tidyr)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyverse)
library(lubridate)
library(viridis)
library(plotly)

df1 <- read_delim("_book/dailyPrecip.csv", delim = "|")

data <- df1 %>%
  rename(Year = "Water Year")

data_long <- data %>%
  pivot_longer(cols = -c(Year, Day),
               names_to = "Month",
               values_to = "Precip")

data_long <- data_long %>%
  mutate(
    MonthNum = match(Month, month.abb),
    Date = make_date(Year, MonthNum, Day),
    MonthNum = NULL
  ) %>%
  mutate(Date = ymd(Date))

# Create a function to reorder months from Oct to Sep
reorder_month <- function(x) {
  factor(x, levels = c("Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep"))
}

# Prepare the data for plotting
data_plot <- data_long %>%
  mutate(
    WaterYear = ifelse(month(Date) >= 10, Year, Year - 1),
    Month = reorder_month(Month)
  ) %>%
  filter(Day == days_in_month(Date)) %>%
  arrange(WaterYear, Month)

# Calculate average cumulative precipitation
avg_precip <- data_plot %>%
  group_by(Month) %>%
  summarize(AvgPrecip = mean(Precip, na.rm = TRUE))
  
data_plot$WaterYear <- as.factor(data_plot$WaterYear)

# Create the plot
p <- ggplot() +
  geom_line(data = data_plot, aes(x = Month, y = Precip, group = WaterYear, color = WaterYear), 
            linewidth = .5) +
  geom_line(data = avg_precip, aes(x = Month, y = AvgPrecip, group = 1), 
            linewidth = 1.5, color = "black") +
  geom_point(data = avg_precip, aes(x = Month, y = AvgPrecip), 
             size = 1, color = "black") +
  scale_x_discrete(limits = levels(reorder_month(avg_precip$Month))) +
  scale_color_viridis(discrete = TRUE, option = "plasma", name = "Water Year") +
  labs(
    title = "Cumulative Precipitation of Water Years 2004-2024",
    x = "Month",
    y = "Cumulative Precipitation (Inches)"
  ) +
  theme_minimal() +
  theme(
   legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.key = element_rect(fill = "white", colour = NA),
    legend.key.size = unit(0.8, "cm")
  )

plotly_plot <- ggplotly(p)

plotly_plot

```

## Growing Season
One of the most striking features of this climate is the dramatic swing in daylight hours throughout the year. Summers bring the phenomenon of the midnight sun, with 24 hours of continuous daylight, while winters plunge the region into weeks of polar night. This extreme light regime profoundly impacts biological rhythms and human activities alike. The coastal plain is also known for its windy conditions, with strong easterly winds being common. These winds, combined with the already frigid temperatures, can create dangerously low wind chill factors.

The growing season in the Arctic Coastal Plain is exceptionally brief, usually lasting only 50-60 days. This short window of relatively warmer temperatures and thawed ground surface supports a unique but fragile tundra ecosystem. Coastal areas are further influenced by sea ice, which is typically present for 8-9 months of the year, affecting local weather patterns and wildlife migrations.

```{r, fig.cap = "This is NRCS SNOTEL data from https://wcc.sc.egov.usda.gov/nwcc/view " , echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(plotly)

# Read the CSV file
solar_data <- read_delim("_book/Solar_Data.txt", 
                         delim = "\t",
                         col_names = c("Date", "SR"),
                         col_types = cols(Date = col_character(), SR = col_double()))

# Process data and remove negative values
solar_data <- solar_data %>%
  mutate(Date = mdy(Date),
         Year = year(Date),
         DayOfYear = yday(Date)) %>%
  filter(SR >= 0)  # Remove negative SR values

# Calculate average SR for each day of the year
average_data <- solar_data %>%
  group_by(DayOfYear) %>%
  summarize(AvgSR = mean(SR, na.rm = TRUE)) %>%
  mutate(Year = "Average")

# Create the plot
p1 <- ggplot() +
  geom_line(data = solar_data, aes(x = DayOfYear, y = SR, color = factor(Year), group = Year), alpha = 0.7) +
  geom_line(data = average_data, aes(x = DayOfYear, y = AvgSR, group = 1), color = "black", size = 1.2) +
  scale_x_continuous(breaks = yday(as.Date(paste0("2000-", 1:12, "-01"))),
                     labels = month.abb,
                     limits = c(1, 366)) +
  scale_color_viridis_d() +
  labs(title = "Daily Solar Radiation 2011-2024 with Average",
       x = "Month",
       y = "Solar Radiation (Watt)",
       color = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

# Convert to interactive plotly plot
plotly_plot <- ggplotly(p1)

# Display the plot
plotly_plot
```

Underlying this harsh surface climate is a layer of continuous permafrost, a defining characteristic of the region. The active layer, which thaws seasonally, is typically shallow, extending only 30-50 cm (12-20 inches) deep. This frozen ground significantly influences the area's ecology and presents unique challenges for construction and resource extraction.

## Geomorphology

The Arctic Coastal Plain is characterized by extremely low relief, with elevations ranging from approximately 4 meters to 94 meters and a mean elevation of 33 meters above sea level. This remarkably flat terrain facilitates the development of several distinctive periglacial and aeolian landforms:

```{r, fig.cap = "This is IFSAR data from https://elevation.alaska.gov/ " , echo = FALSE, warning=FALSE, message=FALSE}

library(ggplot2)
library(tidyr)
library(dplyr)
library(plotly)

# Create the data frame
data <- data.frame(
  Label = c("4.68-7.48", "7.48-10.28", "10.28-13.08", "13.08-15.88", "15.88-18.68", 
            "18.68-21.48", "21.48-24.28", "24.28-27.08", "27.08-29.88", "29.88-32.68",
            "32.68-35.48", "35.48-38.28", "38.28-41.08", "41.08-43.88", "43.88-46.68",
            "46.68-49.48", "49.48-52.28", "52.28-55.08", "55.08-57.88", "57.88-60.68",
            "60.68-63.48", "63.48-66.28", "66.28-69.08", "69.08-71.88", "71.88-74.67",
            "74.67-77.47", "77.47-80.27", "80.27-83.07", "83.07-85.87", "85.87-88.67",
            "88.67-91.47", "91.47-94.27"),
  Count = c(349170, 1094535, 1309892, 2263142, 3097026, 4285123, 7421257, 8894325,
            9168480, 9663489, 8460041, 7194482, 5050500, 4227242, 3466091, 3207795,
            2481361, 1641636, 1191998, 752523, 543485, 457052, 335404, 434094,
            201831, 158215, 84838, 57487, 59461, 48412, 14911, 5396)
)

# Extract middle values of elevation ranges for x-axis
data$mid_elevation <- sapply(strsplit(as.character(data$Label), "-"), function(x) {
  mean(as.numeric(x))
})

# Calculate simple mean
mean_elevation <- 33

# Create the histogram
p4 <- ggplot(data, aes(x = mid_elevation, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_vline(aes(xintercept = mean_elevation), 
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Elevation Distribution of the Arctic Coastal Plain",
       x = "Elevation (meters)",
       y = "Count",
       caption = paste("Mean elevation:", round(mean_elevation, 2), "meters")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(0, 95, by = 10))

plotly_plot <- ggplotly(p4)

plotly_plot

```


Longitudinal Dunes:

Ancient sand dune systems oriented parallel to prevailing wind directions
Formed during past climatic conditions when the region was less vegetated
Now mostly stabilized by tundra vegetation but still influence modern topography

Oriented Lakes (Thermokarst Lakes):

Distinctive oval-shaped lakes aligned perpendicular to prevailing summer winds
Formation controlled by differential erosion and thaw along lake margins
Typically 2-4 meters deep with long axes oriented 10-15 degrees west of true north
Result from permafrost thaw and wind-driven circulation patterns

Highly Sinuous River Channels:

Extremely meandering channels due to the very low gradient coastal plain
Channels often confined by ice-rich permafrost banks
Seasonal flow regimes heavily influenced by spring breakup events

Ice Wedge Polygons:

Extensive polygonal ground patterns formed by thermal contraction cracking
Create distinctive micro-topography across the plain
Range from low-centered to high-centered depending on degradation state
Typically 10-30 meters in diameter

Beaded Streams:

Distinctive "string of beads" appearance from connected pools
Formed where channels intersect ice wedges
Pools typically 2-3 meters deep with connecting channels
Important habitat for fish and waterbirds

This landscape represents a complex interaction between periglacial, fluvial, and aeolian processes operating on extremely ice-rich permafrost terrain. The low relief makes the region particularly sensitive to climate change, with thermokarst processes actively modifying the surface morphology.
Landscape/Landforms: Insert landscape diagram of area

