# General Nature of the Area

## Climate

The Arctic Coastal Plain of Alaska is characterized by a harsh tundra climate, classified as ET in the Köppen system. This region experiences extreme temperature variations, with brutally cold winters where average temperatures often plummet below -22°F, and brief, cool summers barely reaching 37-54°F in July. The average annual temperature ranges from 8 to 14°F. The average freeze-free period is fewer than 5 days to 20 days. Freezing temperatures can occur in any month.

## Insert 30 year normal temperture plot and or potentially an interactive plot that shows every year with the 30 year normal on top

Despite its frigid nature, the coastal plain receives surprisingly little precipitation, typically ranging from 4-10 inches annually. Most of this falls as snow, blanketing the landscape for much of the year. The average annual snowfall is about 50 to 100 centimeters. The low precipitation, coupled with minimal evaporation rates due to the cold, creates a unique hydrologic balance that doesn't quite fit the traditional definition of a desert.

## Insert 30 year normal precipitation plot 

One of the most striking features of this climate is the dramatic swing in daylight hours throughout the year. Summers bring the phenomenon of the midnight sun, with 24 hours of continuous daylight, while winters plunge the region into weeks of polar night. This extreme light regime profoundly impacts biological rhythms and human activities alike. The coastal plain is also known for its windy conditions, with strong easterly winds being common. These winds, combined with the already frigid temperatures, can create dangerously low wind chill factors.

The growing season in the Arctic Coastal Plain is exceptionally brief, usually lasting only 50-60 days. This short window of relatively warmer temperatures and thawed ground supports a unique but fragile tundra ecosystem. Coastal areas are further influenced by sea ice, which is typically present for 8-9 months of the year, affecting local weather patterns and wildlife migrations.

Underlying this harsh surface climate is a layer of continuous permafrost, a defining characteristic of the region. The active layer, which thaws seasonally, is typically shallow, extending only 30-50 cm (12-20 inches) deep. This frozen ground significantly influences the area's ecology and presents unique challenges for construction and resource extraction.

1. Label the heading: `# Hello world {#nice-label}`. 
    - Leave the label off if you like the automated heading generated based on your heading title: for example, `# Hello world` = `# Hello world {#hello-world}`.
    - To label an un-numbered heading, use: `# Hello world {-#nice-label}` or `{# Hello world .unnumbered}`.

1. Next, reference the labeled heading anywhere in the text using `\@ref(nice-label)`; for example, please see Chapter \@ref(cross). 
    - If you prefer text as the link instead of a numbered reference use: [any text you want can go here](#cross).

## Captioned figures and tables

Figures and tables *with captions* can also be cross-referenced from elsewhere in your book using `\@ref(fig:chunk-label)` and `\@ref(tab:chunk-label)`, respectively.

See Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Don't miss Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(pressure, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```
